---
title: "Pre-filtered Majority Voting versus Standard Stacking (Spambase)"
author: "Thanos"
date: "2026-07-12"
output: html_document
params:
  split_seed: 700
  configuration_id: "SpambasePrefilter5of6OOFThresholds"  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose of this file

This file compares four decision-fusion strategies on exactly the same train,
validation, test split and the same six heterogeneous base learners:

1. standalone XGBoost;
2. hard voting;
3. the current XGBoost stacked ensemble;
4. a paper-inspired pre-filtering scheme.

The paper uses seven homogeneous DNNs and accepts a vote directly when at least
`n - 1` learners agree. Only the remaining difficult instances are sent to an
XGBoost meta-learner. Here the same rule is adapted to the six existing base
learners, so an instance bypasses the meta-learner when at least five of the six
base learners predict the same class. This isolates the effect of the fusion
rule; it does not reproduce the paper's DNN architecture or its delete-one-seventh
bootstrap construction.

The base votes use a probability threshold of 0.5, matching the paper's use of
the class with maximum predicted probability. Model-selection thresholds for
standalone XGBoost and the meta-learners are still chosen on the validation set
using balanced accuracy, because the current dataset is imbalanced.

## Libraries


```{r}
library(earth)
library(caret)
library(xgboost)
library(e1071)
library(glmnet)
library(ranger)
library(mgcv)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Load and prepare the data

```{r}
spambase_columns <- c(
  "word_freq_make", "word_freq_address", "word_freq_all", "word_freq_3d",
  "word_freq_our", "word_freq_over", "word_freq_remove", "word_freq_internet",
  "word_freq_order", "word_freq_mail", "word_freq_receive", "word_freq_will",
  "word_freq_people", "word_freq_report", "word_freq_addresses", "word_freq_free",
  "word_freq_business", "word_freq_email", "word_freq_you", "word_freq_credit",
  "word_freq_your", "word_freq_font", "word_freq_000", "word_freq_money",
  "word_freq_hp", "word_freq_hpl", "word_freq_george", "word_freq_650",
  "word_freq_lab", "word_freq_labs", "word_freq_telnet", "word_freq_857",
  "word_freq_data", "word_freq_415", "word_freq_85", "word_freq_technology",
  "word_freq_1999", "word_freq_parts", "word_freq_pm", "word_freq_direct",
  "word_freq_cs", "word_freq_meeting", "word_freq_original", "word_freq_project",
  "word_freq_re", "word_freq_edu", "word_freq_table", "word_freq_conference",
  "char_freq_semicolon", "char_freq_paren", "char_freq_bracket",
  "char_freq_bang", "char_freq_dollar", "char_freq_hash",
  "capital_run_length_average", "capital_run_length_longest",
  "capital_run_length_total", "spam"
)

df <- read.csv(
  "spambase.data",
  header = FALSE,
  col.names = spambase_columns
)

df$spam <- factor(
  df$spam,
  levels = c(0, 1)
)
```

## Train, validation, and test split

```{r}
split_seed <- params$split_seed
configuration_id <- params$configuration_id

set.seed(split_seed)

train_validation_index <- createDataPartition(
  df$spam,
  p = 0.8,
  list = FALSE
)

train_validation <- df[train_validation_index, ]
test <- df[-train_validation_index, ]

train_index <- createDataPartition(
  train_validation$spam,
  p = 0.8,
  list = FALSE
)

train <- train_validation[train_index, ]
validation <- train_validation[-train_index, ]
```

## Evaluation functions

```{r}
evaluate_model <- function(predicted, actual) {
  
  predicted <- factor(predicted, levels = c(0, 1))
  actual <- factor(actual, levels = c(0, 1))
  
  cm <- table(
    Predicted = predicted,
    Actual = actual
  )
  
  TN <- cm["0", "0"]
  FN <- cm["0", "1"]
  FP <- cm["1", "0"]
  TP <- cm["1", "1"]
  
  accuracy <- (TP + TN) / sum(cm)
  
  precision <- ifelse(
    TP + FP == 0,
    0,
    TP / (TP + FP)
  )
  
  recall <- ifelse(
    TP + FN == 0,
    0,
    TP / (TP + FN)
  )
  
  specificity <- ifelse(
    TN + FP == 0,
    0,
    TN / (TN + FP)
  )
  
  f1 <- ifelse(
    precision + recall == 0,
    0,
    2 * precision * recall /
      (precision + recall)
  )
  
  balanced_accuracy <- (
    recall + specificity
  ) / 2
  
  c(
    Accuracy = accuracy,
    Balanced_Accuracy = balanced_accuracy,
    Precision = precision,
    Recall = recall,
    Specificity = specificity,
    F1 = f1
  )
}
```

```{r}
find_threshold <- function(probabilities,
                           actual,
                           metric = "Balanced_Accuracy") {
  
  thresholds <- seq(
    0.05,
    0.95,
    by = 0.01
  )
  
  scores <- sapply(
    thresholds,
    function(threshold) {
      
      predicted <- ifelse(
        probabilities >= threshold,
        "1",
        "0"
      )
      
      predicted <- factor(
        predicted,
        levels = c("0", "1")
      )
      
      evaluate_model(
        predicted,
        actual
      )[metric]
    }
  )
  
  thresholds[which.max(scores)]
}
```

## Voting and pre-filtering functions

```{r}
make_vote_information <- function(probability_features,
                                  probability_threshold = 0.5,
                                  confidence_votes = NULL) {
  
  probability_matrix <- as.matrix(probability_features)
  
  if (!is.numeric(probability_matrix)) {
    stop("All base-learner features must be numeric probabilities.")
  }
  
  number_of_models <- ncol(probability_matrix)
  
  if (is.null(confidence_votes)) {
    confidence_votes <- number_of_models - 1L
  }
  
  if (length(probability_threshold) == 1L) {
    probability_threshold <- rep(
      probability_threshold,
      number_of_models
    )
  }
  
  if (length(probability_threshold) != number_of_models) {
    stop(
      paste(
        "probability_threshold must contain either one value",
        "or one value per base learner."
      )
    )
  }
  
  if (
    confidence_votes < 1L ||
    confidence_votes > number_of_models
  ) {
    stop("confidence_votes must be between 1 and the ensemble size.")
  }
  
  hard_predictions <- matrix(
    FALSE,
    nrow = nrow(probability_matrix),
    ncol = number_of_models
  )
  
  for (model_number in seq_len(number_of_models)) {
    hard_predictions[, model_number] <-
      probability_matrix[, model_number] >=
      probability_threshold[model_number]
  }
  
  colnames(hard_predictions) <- colnames(
    probability_matrix
  )
  
  positive_votes <- rowSums(hard_predictions)
  negative_votes <- number_of_models - positive_votes
  maximum_votes <- pmax(positive_votes, negative_votes)
  
  majority_label <- ifelse(
    positive_votes > negative_votes,
    "1",
    ifelse(
      negative_votes > positive_votes,
      "0",
      NA_character_
    )
  )
  
  confident <- maximum_votes >= confidence_votes
  
  list(
    summary = data.frame(
      Positive_Votes = positive_votes,
      Negative_Votes = negative_votes,
      Maximum_Votes = maximum_votes,
      Majority_Label = majority_label,
      Confident = confident,
      stringsAsFactors = FALSE
    ),
    hard_predictions = hard_predictions,
    probability_thresholds = probability_threshold,
    confidence_votes = confidence_votes
  )
}
```

```{r}
compose_prefilter_predictions <- function(meta_probabilities,
                                          vote_information,
                                          meta_threshold) {
  
  vote_summary <- vote_information$summary
  
  if (length(meta_probabilities) != nrow(vote_summary)) {
    stop(
      paste(
        "meta_probabilities and vote_information",
        "must refer to the same observations."
      )
    )
  }
  
  final_labels <- ifelse(
    meta_probabilities >= meta_threshold,
    "1",
    "0"
  )
  
  confident_index <- which(vote_summary$Confident)
  
  if (length(confident_index) > 0L) {
    final_labels[confident_index] <-
      vote_summary$Majority_Label[confident_index]
  }
  
  factor(
    final_labels,
    levels = c("0", "1")
  )
}
```

```{r}
find_prefilter_threshold <- function(meta_probabilities,
                                     vote_information,
                                     actual,
                                     metric = "Balanced_Accuracy") {
  
  thresholds <- seq(
    0.05,
    0.95,
    by = 0.01
  )
  
  scores <- sapply(
    thresholds,
    function(threshold) {
      
      predicted <- compose_prefilter_predictions(
        meta_probabilities = meta_probabilities,
        vote_information = vote_information,
        meta_threshold = threshold
      )
      
      evaluate_model(
        predicted,
        actual
      )[metric]
    }
  )
  
  thresholds[which.max(scores)]
}
```

```{r}
make_hard_vote_predictions <- function(
    probability_features,
    probability_threshold = 0.5) {
  
  probability_matrix <- as.matrix(
    probability_features
  )
  
  number_of_models <- ncol(
    probability_matrix
  )
  
  if (length(probability_threshold) == 1L) {
    probability_threshold <- rep(
      probability_threshold,
      number_of_models
    )
  }
  
  if (
    length(probability_threshold) !=
      number_of_models
  ) {
    stop(
      paste(
        "probability_threshold must contain",
        "one value or one threshold per learner."
      )
    )
  }
  
  vote_information <- make_vote_information(
    probability_features =
      probability_features,
    probability_threshold =
      probability_threshold,
    confidence_votes = 1L
  )
  
  vote_summary <- vote_information$summary
  final_labels <- vote_summary$Majority_Label
  
  # A six-model ensemble may produce a 3:3 tie.
  # Resolve it using the average signed distance
  # from each learner's own threshold.
  tied <- is.na(final_labels)
  
  if (any(tied)) {
    
    probability_margins <- sweep(
      probability_matrix,
      MARGIN = 2,
      STATS = probability_threshold,
      FUN = "-"
    )
    
    mean_margin <- rowMeans(
      probability_margins
    )
    
    final_labels[tied] <- ifelse(
      mean_margin[tied] >= 0,
      "1",
      "0"
    )
  }
  
  factor(
    final_labels,
    levels = c("0", "1")
  )
}
```

# Standalone XGBoost

## Prepare XGBoost data

```{r}
x_train <- model.matrix(
  spam ~ . - 1,
  data = train
)

x_validation <- model.matrix(
  spam ~ . - 1,
  data = validation
)

x_test <- model.matrix(
  spam ~ . - 1,
  data = test
)

y_train <- as.numeric(train$spam) - 1
y_validation <- as.numeric(validation$spam) - 1
y_test <- as.numeric(test$spam) - 1

dtrain <- xgb.DMatrix(
  data = x_train,
  label = y_train
)

dvalidation <- xgb.DMatrix(
  data = x_validation,
  label = y_validation
)

dtest <- xgb.DMatrix(
  data = x_test,
  label = y_test
)
```

## Calculate class weight

```{r}
xgb_positive_weight <- sum(y_train == 0) /
  sum(y_train == 1)

xgb_positive_weight
```

## Train standalone XGBoost

```{r}
set.seed(100)

xgb_parameters <- list(
  objective = "binary:logistic",
  eval_metric = "aucpr",
  tree_method = "hist",
  max_depth = 3,
  learning_rate = 0.05,
  min_child_weight = 5,
  subsample = 0.8,
  colsample_bytree = 0.8,
  reg_lambda = 1,
  reg_alpha = 0,
  scale_pos_weight = xgb_positive_weight,
  seed = 100
)

xgb_model <- xgb.train(
  params = xgb_parameters,
  data = dtrain,
  nrounds = 1000,
  evals = list(
    train = dtrain,
    validation = dvalidation
  ),
  early_stopping_rounds = 30,
  maximize = TRUE,
  verbose = 1,
  print_every_n = 25
)
```

## Select standalone XGBoost threshold

```{r}
xgb_validation_prob <- predict(
  xgb_model,
  dvalidation
)

xgb_threshold <- find_threshold(
  probabilities = xgb_validation_prob,
  actual = validation$spam,
  metric = "Balanced_Accuracy"
)

xgb_threshold
```

## Evaluate standalone XGBoost

```{r}
xgb_prob <- predict(
  xgb_model,
  dtest
)

xgb_pred <- factor(
  ifelse(
    xgb_prob >= xgb_threshold,
    "1",
    "0"
  ),
  levels = c("0", "1")
)

xgb_metrics <- evaluate_model(
  xgb_pred,
  test$spam
)

xgb_metrics
```

# Base learners for stacking

## Elastic-net logistic regression

```{r}
# Convert data frames to numeric predictor matrices
x_train <- model.matrix(
  spam ~ . - 1,
  data = train
)

x_validation <- model.matrix(
  spam ~ . - 1,
  data = validation
)

x_test <- model.matrix(
  spam ~ . - 1,
  data = test
)

# Convert factor outcome from levels "0"/"1" to numeric 0/1
y_train <- as.numeric(train$spam) - 1

set.seed(500)

elastic_net_model <- cv.glmnet(
  x = x_train,
  y = y_train,
  family = "binomial",
  alpha = 0.25,
  type.measure = "deviance",
  nfolds = 5
)

elastic_net_validation_prob <- as.numeric(
  predict(
    elastic_net_model,
    newx = x_validation,
    s = "lambda.min",
    type = "response"
  )
)

elastic_net_test_prob <- as.numeric(
  predict(
    elastic_net_model,
    newx = x_test,
    s = "lambda.min",
    type = "response"
  )
)
```

## Generalized additive model

```{r}
predictor_names <- setdiff(
  names(train),
  "spam"
)

gam_formula <- as.formula(
  paste(
    "spam ~",
    paste(
      sprintf(
        "s(`%s`, k = 5)",
        predictor_names
      ),
      collapse = " + "
    )
  )
)

set.seed(550)

gam_model <- gam(
  formula = gam_formula,
  data = train,
  family = binomial,
  method = "REML",
  select = TRUE
)

gam_validation_prob <- as.numeric(
  predict(
    gam_model,
    newdata = validation,
    type = "response"
  )
)

gam_test_prob <- as.numeric(
  predict(
    gam_model,
    newdata = test,
    type = "response"
  )
)
```

## Random forest

```{r}
set.seed(600)

rf_positive_weight <-
  sum(train$spam == "0") /
  sum(train$spam == "1")

rf_model <- ranger(
  spam ~ .,
  data = train,
  probability = TRUE,
  num.trees = 1000,
  mtry = 3,
  min.node.size = 10,
  sample.fraction = 0.8,
  class.weights = c(
    "0" = 1,
    "1" = rf_positive_weight
  ),
  seed = 600
)

rf_validation_prob <- predict(
  rf_model,
  data = validation
)$predictions[, "1"]

rf_test_prob <- predict(
  rf_model,
  data = test
)$predictions[, "1"]
```

## Tuned radial support vector machine

```{r}
#svm_positive_weight <-
  #sum(train$spam == "0") /
  #sum(train$spam == "1")

set.seed(700)

svm_tuning <- tune.svm(
  spam ~ .,
  data = train,
  kernel = "radial",
  probability = TRUE,
  scale = TRUE,
  cost = c(
    0.1,
    1,
    10
  ),
  gamma = c(
    0.01,
    0.05,
    0.1
  ),
  #class.weights = c(
    #"0" = 1,
    #"1" = svm_positive_weight
  #),
  tunecontrol = tune.control(
    sampling = "cross",
    cross = 3
  )
)

svm_model <- svm_tuning$best.model

svm_tuning$best.parameters

svm_validation_raw <- predict(
  svm_model,
  newdata = validation,
  probability = TRUE
)

svm_validation_prob <- attr(
  svm_validation_raw,
  "probabilities"
)[, "1"]

svm_test_raw <- predict(
  svm_model,
  newdata = test,
  probability = TRUE
)

svm_test_prob <- attr(
  svm_test_raw,
  "probabilities"
)[, "1"]
```

## Multivariate adaptive regression spline

```{r}
set.seed(800)

mars_model <- earth(
  spam ~ .,
  data = train,
  degree = 2,
  nprune = 20,
  glm = list(
    family = binomial
  )
)

mars_validation_prob <- as.numeric(
  predict(
    mars_model,
    newdata = validation,
    type = "response"
  )
)

mars_test_prob <- as.numeric(
  predict(
    mars_model,
    newdata = test,
    type = "response"
  )
)
```


# Out-of-fold predictions for the meta-learner

```{r}
set.seed(300)

stack_folds <- createFolds(
  train$spam,
  k = 5,
  list = TRUE,
  returnTrain = FALSE
)

stack_train_features <- data.frame(
  ElasticNet = rep(NA_real_, nrow(train)),
  GAM = rep(NA_real_, nrow(train)),
  MARS = rep(NA_real_, nrow(train)),
  RandomForest = rep(NA_real_, nrow(train)),
  XGBoost = rep(NA_real_, nrow(train)),
  SVM = rep(NA_real_, nrow(train))
)

best_iteration <- xgb.attr(
  xgb_model,
  "best_iteration"
)

if (is.null(best_iteration)) {
  stop(
    paste(
      "The standalone XGBoost model does not contain",
      "a best_iteration attribute.",
      "Run the XGBoost training block first."
    )
  )
}

xgb_stack_nrounds <- as.integer(
  best_iteration
) + 1L

if (
  length(xgb_stack_nrounds) != 1 ||
  is.na(xgb_stack_nrounds) ||
  xgb_stack_nrounds < 1
) {
  stop("xgb_stack_nrounds is not a valid positive integer.")
}

xgb_stack_nrounds
```

```{r}

for (fold_number in seq_along(stack_folds)) {
  
  fold_validation_index <- stack_folds[[fold_number]]
  
  fold_train <- train[-fold_validation_index, ]
  fold_validation <- train[fold_validation_index, ]
  
  # Elastic-net logistic regression

  fold_elastic_x_train <- model.matrix(
    spam ~ . - 1,
    data = fold_train
  )

  fold_elastic_x_validation <- model.matrix(
    spam ~ . - 1,
    data = fold_validation
  )

  fold_elastic_y_train <-
    as.numeric(fold_train$spam) - 1

  set.seed(500 + fold_number)

  fold_elastic_net_model <- cv.glmnet(
    x = fold_elastic_x_train,
    y = fold_elastic_y_train,
    family = "binomial",
    alpha = 0.25,
    type.measure = "deviance",
    nfolds = 5
  )

  stack_train_features$ElasticNet[
    fold_validation_index
  ] <- as.numeric(
    predict(
      fold_elastic_net_model,
      newx = fold_elastic_x_validation,
      s = "lambda.min",
      type = "response"
    )
  )
  
  # Generalized additive model

  set.seed(550 + fold_number)

  fold_gam_model <- gam(
    formula = gam_formula,
    data = fold_train,
    family = binomial,
    method = "REML",
    select = TRUE
  )

  stack_train_features$GAM[
    fold_validation_index
  ] <- as.numeric(
    predict(
      fold_gam_model,
      newdata = fold_validation,
      type = "response"
    )
  )
  
  # Multivariate adaptive regression spline
  set.seed(800 + fold_number)

  fold_mars_model <- earth(
    spam ~ .,
    data = fold_train,
    degree = 2,
    nprune = 20,
    glm = list(
      family = binomial
    )
  )

  stack_train_features$MARS[
    fold_validation_index
  ] <- as.numeric(
    predict(
      fold_mars_model,
      newdata = fold_validation,
      type = "response"
    )
  )
  
  # Random forest
  set.seed(300 + fold_number)

  fold_positive_weight <-
    sum(fold_train$spam == "0") /
    sum(fold_train$spam == "1")

  fold_rf_model <- ranger(
    spam ~ .,
    data = fold_train,
    probability = TRUE,
    num.trees = 1000,
    mtry = 3,
    min.node.size = 10,
    sample.fraction = 0.8,
    class.weights = c(
      "0" = 1,
      "1" = fold_positive_weight
    ),
    seed = 300 + fold_number
  )

  stack_train_features$RandomForest[
    fold_validation_index
  ] <- predict(
    fold_rf_model,
    data = fold_validation
  )$predictions[, "1"]
  
  # XGBoost
  fold_x_train <- model.matrix(
    spam ~ . - 1,
    data = fold_train
  )
  
  fold_x_validation <- model.matrix(
    spam ~ . - 1,
    data = fold_validation
  )
  
  fold_y_train <- as.numeric(
    fold_train$spam
  ) - 1
  
  fold_dtrain <- xgb.DMatrix(
    data = fold_x_train,
    label = fold_y_train
  )
  
  fold_dvalidation <- xgb.DMatrix(
    data = fold_x_validation
  )
  
  fold_xgb_parameters <- xgb_parameters
  fold_xgb_parameters$scale_pos_weight <-
    sum(fold_y_train == 0) /
    sum(fold_y_train == 1)
  fold_xgb_parameters$seed <- 300 + fold_number
  
  fold_xgb_model <- xgb.train(
    params = fold_xgb_parameters,
    data = fold_dtrain,
    nrounds = xgb_stack_nrounds,
    verbose = 0
  )
  
  stack_train_features$XGBoost[
    fold_validation_index
  ] <- predict(
    fold_xgb_model,
    fold_dvalidation
  )
  
  # Tuned radial SVM

  #fold_svm_positive_weight <-
    #sum(fold_train$spam == "0") /
    #sum(fold_train$spam == "1")

  set.seed(700 + fold_number)

  fold_svm_tuning <- tune.svm(
    spam ~ .,
    data = fold_train,
    kernel = "radial",
    probability = TRUE,
    scale = TRUE,
    cost = c(
      0.1,
      1,
      10
    ),
    gamma = c(
      0.01,
      0.05,
      0.1
    ),
    #class.weights = c(
      #"0" = 1,
      #"1" = fold_svm_positive_weight
    #),
    tunecontrol = tune.control(
      sampling = "cross",
      cross = 3
    )
  )

  fold_svm_model <-
    fold_svm_tuning$best.model

  fold_svm_raw <- predict(
    fold_svm_model,
    newdata = fold_validation,
    probability = TRUE
  )

  stack_train_features$SVM[
    fold_validation_index
  ] <- attr(
    fold_svm_raw,
    "probabilities"
  )[, "1"]
}
```

# Inspect the out-of-fold probability features
```{r}
summary(stack_train_features)

round(
  cor(stack_train_features),
  digits = 3
)
```


# Evaluate the out-of-fold base learners
```{r}
oof_base_results <- lapply(
  names(stack_train_features),
  function(model_name) {
    
    probabilities <-
      stack_train_features[[model_name]]
    
    threshold <- find_threshold(
      probabilities = probabilities,
      actual = train$spam,
      metric = "Balanced_Accuracy"
    )
    
    predictions <- factor(
      ifelse(
        probabilities >= threshold,
        "1",
        "0"
      ),
      levels = c("0", "1")
    )
    
    metrics <- evaluate_model(
      predictions,
      train$spam
    )
    
    data.frame(
      Model = model_name,
      Threshold = threshold,
      t(metrics),
      row.names = NULL
    )
  }
)

oof_base_results <- do.call(
  rbind,
  oof_base_results
)

oof_base_results
```

# XGBoost meta-learner

## Construct meta-learning datasets

```{r}
stack_validation_features <- data.frame(
  ElasticNet = as.numeric(
    elastic_net_validation_prob
  ),
  GAM = as.numeric(
    gam_validation_prob
  ),
  MARS = as.numeric(
    mars_validation_prob
  ),
  RandomForest = as.numeric(
    rf_validation_prob
  ),
  XGBoost = as.numeric(
    xgb_validation_prob
  ),
  SVM = as.numeric(
    svm_validation_prob
  )
)

stack_test_features <- data.frame(
  ElasticNet = as.numeric(
    elastic_net_test_prob
  ),
  GAM = as.numeric(
    gam_test_prob
  ),
  MARS = as.numeric(
    mars_test_prob
  ),
  RandomForest = as.numeric(
    rf_test_prob
  ),
  XGBoost = as.numeric(
    xgb_prob
  ),
  SVM = as.numeric(
    svm_test_prob
  )
)

stack_train_labels <- as.numeric(train$spam) - 1
stack_validation_labels <- as.numeric(validation$spam) - 1

stack_positive_weight <- sum(stack_train_labels == 0) /
  sum(stack_train_labels == 1)

stopifnot(
  !anyNA(stack_train_features),
  !anyNA(stack_validation_features),
  !anyNA(stack_test_features)
)

stopifnot(
  identical(
    colnames(stack_train_features),
    colnames(stack_validation_features)
  ),
  identical(
    colnames(stack_train_features),
    colnames(stack_test_features)
  )
)

colnames(stack_train_features)

dstack_train <- xgb.DMatrix(
  data = as.matrix(stack_train_features),
  label = stack_train_labels
)

dstack_validation <- xgb.DMatrix(
  data = as.matrix(stack_validation_features),
  label = stack_validation_labels
)

dstack_test <- xgb.DMatrix(
  data = as.matrix(stack_test_features)
)

```
Individual base-learner performance on the test set

```{r}
stopifnot(
  identical(
    colnames(stack_validation_features),
    colnames(stack_test_features)
  )
)

base_learner_test_results <- lapply(
  colnames(stack_validation_features),
  function(model_name) {
    
    validation_probabilities <-
      stack_validation_features[[model_name]]
    
    test_probabilities <-
      stack_test_features[[model_name]]
    
    # Select the threshold using validation data only
    selected_threshold <- find_threshold(
      probabilities = validation_probabilities,
      actual = validation$spam,
      metric = "Balanced_Accuracy"
    )
    
    # Apply the validation-selected threshold to the test data
    test_predictions <- factor(
      ifelse(
        test_probabilities >= selected_threshold,
        "1",
        "0"
      ),
      levels = c("0", "1")
    )
    
    test_metrics <- evaluate_model(
      predicted = test_predictions,
      actual = test$spam
    )
    
    data.frame(
      Seed = split_seed,
      Configuration_ID = configuration_id,
      Model = model_name,
      Threshold = selected_threshold,
      Accuracy = unname(
        test_metrics["Accuracy"]
      ),
      Balanced_Accuracy = unname(
        test_metrics["Balanced_Accuracy"]
      ),
      Precision = unname(
        test_metrics["Precision"]
      ),
      Recall = unname(
        test_metrics["Recall"]
      ),
      Specificity = unname(
        test_metrics["Specificity"]
      ),
      F1 = unname(
        test_metrics["F1"]
      ),
      row.names = NULL
    )
  }
)

# Convert the list returned by lapply into one data frame
base_learner_test_results <- do.call(
  rbind,
  base_learner_test_results
)

# Convert the six model rows into one wide row for the final CSV
base_learner_result_row <- base_learner_test_results |>
  select(
    -Seed,
    -Configuration_ID
  ) |>
  pivot_wider(
    names_from = Model,
    values_from = c(
      Threshold,
      Accuracy,
      Balanced_Accuracy,
      Precision,
      Recall,
      Specificity,
      F1
    ),
    names_glue = "Base_{Model}_{.value}"
  )

base_learner_test_results
```

## Train the XGBoost meta-learner

```{r}
set.seed(500)

stack_parameters <- list(
  objective = "binary:logistic",
  eval_metric = "aucpr",
  tree_method = "hist",
  max_depth = 2,
  learning_rate = 0.05,
  min_child_weight = 5,
  subsample = 0.8,
  colsample_bytree = 1,
  reg_lambda = 1,
  reg_alpha = 0,
  scale_pos_weight = stack_positive_weight,
  seed = 500
)

stack_model <- xgb.train(
  params = stack_parameters,
  data = dstack_train,
  nrounds = 500,
  evals = list(
    train = dstack_train,
    validation = dstack_validation
  ),
  early_stopping_rounds = 30,
  maximize = TRUE,
  verbose = 1,
  print_every_n = 25
)
```

## Inspect meta-learner feature importance

```{r}
stack_importance <- xgb.importance(
  model = stack_model,
  feature_names = colnames(
    stack_train_features
  )
)

stack_importance

if (nrow(stack_importance) > 0) {
  xgb.plot.importance(
    stack_importance
  )
}
```


## Select the ensemble threshold

```{r}
stack_validation_prob <- predict(
  stack_model,
  dstack_validation
)

stack_threshold <- find_threshold(
  probabilities = stack_validation_prob,
  actual = validation$spam,
  metric = "Balanced_Accuracy"
)

stack_threshold
```

## Evaluate the stacked ensemble

```{r}
stack_prob <- predict(
  stack_model,
  dstack_test
)

stack_pred <- factor(
  ifelse(
    stack_prob >= stack_threshold,
    "1",
    "0"
  ),
  levels = c("0", "1")
)

stack_metrics <- evaluate_model(
  stack_pred,
  test$spam
)

stack_metrics
```

# Paper-inspired pre-filtering by strict majority agreement

## Construct hard-vote information

```{r}
number_of_base_learners <- ncol(
  stack_train_features
)

# Select one voting threshold per learner using OOF training predictions
base_vote_thresholds <- vapply(
  colnames(stack_train_features),
  function(model_name) {
    
    find_threshold(
      probabilities =
        stack_train_features[[model_name]],
      actual = train$spam,
      metric = "Balanced_Accuracy"
    )
  },
  numeric(1)
)

# Ensure thresholds follow the exact meta-feature column order
base_vote_thresholds <- base_vote_thresholds[
  colnames(stack_train_features)
]

# Use learner-specific thresholds instead of a common threshold of 0.5
paper_probability_threshold <-
  base_vote_thresholds

# Current adapted experiment: agreement of at least n - 1 learners
# (5 of 6 with the current ensemble size) bypasses XGBoost.
paper_confidence_votes <-
  number_of_base_learners - 1L

base_vote_thresholds

paper_train_votes <- make_vote_information(
  probability_features = stack_train_features,
  probability_threshold = paper_probability_threshold,
  confidence_votes = paper_confidence_votes
)

paper_validation_votes <- make_vote_information(
  probability_features = stack_validation_features,
  probability_threshold = paper_probability_threshold,
  confidence_votes = paper_confidence_votes
)

paper_test_votes <- make_vote_information(
  probability_features = stack_test_features,
  probability_threshold = paper_probability_threshold,
  confidence_votes = paper_confidence_votes
)

c(
  Ensemble_Size = number_of_base_learners,
  Required_Agreement = paper_confidence_votes
)
```

## Inspect how much data bypasses the meta-learner

```{r}
prefilter_diagnostics <- data.frame(
  Data_Set = c(
    "OOF training",
    "Validation",
    "Test"
  ),
  Observations = c(
    nrow(stack_train_features),
    nrow(stack_validation_features),
    nrow(stack_test_features)
  ),
  Direct_Vote_Predictions = c(
    sum(paper_train_votes$summary$Confident),
    sum(paper_validation_votes$summary$Confident),
    sum(paper_test_votes$summary$Confident)
  )
)

prefilter_diagnostics$Sent_To_Meta_Learner <-
  prefilter_diagnostics$Observations -
  prefilter_diagnostics$Direct_Vote_Predictions

prefilter_diagnostics$Direct_Vote_Rate <- round(
  prefilter_diagnostics$Direct_Vote_Predictions /
    prefilter_diagnostics$Observations,
  digits = 3
)

prefilter_diagnostics$Meta_Learner_Rate <- round(
  prefilter_diagnostics$Sent_To_Meta_Learner /
    prefilter_diagnostics$Observations,
  digits = 3
)

prefilter_diagnostics
```

```{r}
# This table checks whether high-agreement cases are almost exclusively
# predictions of the majority class. That would make a high bypass rate less
# informative on an imbalanced dataset.
table(
  Confident = paper_train_votes$summary$Confident,
  Majority_Label = paper_train_votes$summary$Majority_Label,
  useNA = "ifany"
)
```

## Filter the OOF training cases

```{r}
paper_train_difficult <-
  !paper_train_votes$summary$Confident

paper_validation_difficult <-
  !paper_validation_votes$summary$Confident

paper_test_difficult <-
  !paper_test_votes$summary$Confident

paper_filtered_train_features <-
  stack_train_features[
    paper_train_difficult,
    ,
    drop = FALSE
  ]

paper_filtered_train_labels <-
  stack_train_labels[paper_train_difficult]

paper_filtered_validation_features <-
  stack_validation_features[
    paper_validation_difficult,
    ,
    drop = FALSE
  ]

paper_filtered_validation_labels <-
  stack_validation_labels[paper_validation_difficult]

if (nrow(paper_filtered_train_features) < 10L) {
  stop(
    paste(
      "Too few difficult OOF observations remain after pre-filtering.",
      "Use more base-learner diversity or reconsider the pre-filtering rule."
    )
  )
}

if (length(unique(paper_filtered_train_labels)) < 2L) {
  stop(
    paste(
      "The filtered OOF training set contains only one class.",
      "The XGBoost meta-learner cannot be trained."
    )
  )
}

data.frame(
  Filtered_Training_Observations =
    nrow(paper_filtered_train_features),
  Negative_Class =
    sum(paper_filtered_train_labels == 0),
  Positive_Class =
    sum(paper_filtered_train_labels == 1)
)
```

## Train XGBoost only on difficult OOF cases

```{r}
paper_positive_weight <-
  sum(paper_filtered_train_labels == 0) /
  sum(paper_filtered_train_labels == 1)

paper_dtrain <- xgb.DMatrix(
  data = as.matrix(
    paper_filtered_train_features
  ),
  label = paper_filtered_train_labels
)

paper_parameters <- list(
  objective = "binary:logistic",
  eval_metric = "aucpr",
  tree_method = "hist",
  max_depth = 2,
  learning_rate = 0.05,
  min_child_weight = 5,
  subsample = 0.8,
  colsample_bytree = 1,
  reg_lambda = 1,
  reg_alpha = 0,
  scale_pos_weight = paper_positive_weight,
  seed = 900
)

set.seed(900)

can_use_filtered_validation <-
  nrow(paper_filtered_validation_features) >= 10L &&
  length(unique(paper_filtered_validation_labels)) == 2L

if (can_use_filtered_validation) {
  
  paper_dvalidation <- xgb.DMatrix(
    data = as.matrix(
      paper_filtered_validation_features
    ),
    label = paper_filtered_validation_labels
  )
  
  paper_meta_model <- xgb.train(
    params = paper_parameters,
    data = paper_dtrain,
    nrounds = 500,
    evals = list(
      train = paper_dtrain,
      validation = paper_dvalidation
    ),
    early_stopping_rounds = 30,
    maximize = TRUE,
    verbose = 1,
    print_every_n = 25
  )
  
} else {
  
  warning(
    paste(
      "The filtered validation set is too small or contains one class.",
      "The pre-filtered meta-learner is trained for a fixed 250 rounds."
    )
  )
  
  paper_meta_model <- xgb.train(
    params = paper_parameters,
    data = paper_dtrain,
    nrounds = 250,
    verbose = 0
  )
}
```

## Select the hybrid meta-learner threshold

```{r}
paper_validation_meta_prob <- predict(
  paper_meta_model,
  dstack_validation
)

# The threshold is selected using the complete hybrid validation prediction:
# confident cases use their vote, while difficult cases use the meta-learner.
paper_meta_threshold <- find_prefilter_threshold(
  meta_probabilities = paper_validation_meta_prob,
  vote_information = paper_validation_votes,
  actual = validation$spam,
  metric = "Balanced_Accuracy"
)

paper_meta_threshold
```

## Evaluate the pre-filtered ensemble

```{r}
paper_test_meta_prob <- predict(
  paper_meta_model,
  dstack_test
)

paper_prefilter_pred <- compose_prefilter_predictions(
  meta_probabilities = paper_test_meta_prob,
  vote_information = paper_test_votes,
  meta_threshold = paper_meta_threshold
)

paper_prefilter_metrics <- evaluate_model(
  paper_prefilter_pred,
  test$spam
)

paper_prefilter_metrics
```

# Hard-voting reference

```{r}
hard_vote_pred <- make_hard_vote_predictions(
  probability_features = stack_test_features,
  probability_threshold = paper_probability_threshold
)

hard_vote_metrics <- evaluate_model(
  hard_vote_pred,
  test$spam
)

hard_vote_metrics
```

# Complete comparison

```{r}
comparison <- data.frame(
  Model = c(
    "Standalone XGBoost",
    "Hard vote (ties by mean probability)",
    "Standard stack: XGBoost meta-learner",
    "Paper-inspired pre-filter + XGBoost meta-learner"
  ),
  Threshold = c(
    xgb_threshold,
    NA_real_,
    stack_threshold,
    paper_meta_threshold
  ),
  rbind(
    xgb_metrics,
    hard_vote_metrics,
    stack_metrics,
    paper_prefilter_metrics
  )
)

rownames(comparison) <- NULL

numeric_columns <- sapply(
  comparison,
  is.numeric
)

comparison[numeric_columns] <- round(
  comparison[numeric_columns],
  digits = 3
)

comparison
```

## Direct comparison with the current standard stack

```{r}
prefilter_minus_standard_stack <-
  paper_prefilter_metrics["Balanced_Accuracy"] -
  stack_metrics["Balanced_Accuracy"]

prefilter_minus_standalone_xgb <-
  paper_prefilter_metrics["Balanced_Accuracy"] -
  xgb_metrics["Balanced_Accuracy"]

if (prefilter_minus_standard_stack > 0) {
  cat(
    "The paper-inspired pre-filtered ensemble surpassed the standard stack by",
    round(prefilter_minus_standard_stack, 3),
    "in balanced accuracy.\n"
  )
} else if (prefilter_minus_standard_stack < 0) {
  cat(
    "The paper-inspired pre-filtered ensemble was",
    round(abs(prefilter_minus_standard_stack), 3),
    "lower than the standard stack in balanced accuracy.\n"
  )
} else {
  cat(
    "The paper-inspired pre-filtered ensemble and the standard stack",
    "had identical balanced accuracy.\n"
  )
}

if (prefilter_minus_standalone_xgb > 0) {
  cat(
    "It also surpassed standalone XGBoost by",
    round(prefilter_minus_standalone_xgb, 3),
    "in balanced accuracy.\n"
  )
} else if (prefilter_minus_standalone_xgb < 0) {
  cat(
    "It remained",
    round(abs(prefilter_minus_standalone_xgb), 3),
    "below standalone XGBoost in balanced accuracy.\n"
  )
} else {
  cat(
    "It matched standalone XGBoost in balanced accuracy.\n"
  )
}
```

# Save the result for repeated split seeds

```{r}
# ------------------------------------------------------------
# Combine ensemble and base-learner results
# ------------------------------------------------------------

ensemble_seed_result <- data.frame(
  Seed = split_seed,
  Configuration_ID = configuration_id,
  
  XGBoost_BA = unname(
    xgb_metrics["Balanced_Accuracy"]
  ),
  
  Hard_Vote_BA = unname(
    hard_vote_metrics["Balanced_Accuracy"]
  ),
  
  Standard_Stack_BA = unname(
    stack_metrics["Balanced_Accuracy"]
  ),
  
  Prefilter_Stack_BA = unname(
    paper_prefilter_metrics[
      "Balanced_Accuracy"
    ]
  ),
  
  Prefilter_Minus_Standard = unname(
    paper_prefilter_metrics[
      "Balanced_Accuracy"
    ] -
      stack_metrics[
        "Balanced_Accuracy"
      ]
  ),
  
  Prefilter_Minus_XGBoost = unname(
    paper_prefilter_metrics[
      "Balanced_Accuracy"
    ] -
      xgb_metrics[
        "Balanced_Accuracy"
      ]
  ),
  
  Direct_Vote_Rate_Test = mean(
    paper_test_votes$summary$Confident
  ),
  
  Test_Cases_Sent_To_Meta = sum(
    paper_test_difficult
  ),
  
  stringsAsFactors = FALSE
)

# Add all individual base-learner values to the same row
vote_threshold_result_row <- as.data.frame(
  as.list(
    setNames(
      as.numeric(
        base_vote_thresholds
      ),
      paste0(
        "Gate_Threshold_",
        names(base_vote_thresholds)
      )
    )
  ),
  check.names = FALSE
)

seed_result <- cbind(
  ensemble_seed_result,
  base_learner_result_row,
  vote_threshold_result_row
)

# ------------------------------------------------------------
# Save results in configuration_id.csv
# ------------------------------------------------------------

results_directory <- "prefilter_seed_results"

dir.create(
  results_directory,
  showWarnings = FALSE,
  recursive = TRUE
)

result_filename <- file.path(
  results_directory,
  paste0(
    configuration_id,
    ".csv"
  )
)

if (file.exists(result_filename)) {
  
  existing_results <- read.csv(
    result_filename,
    stringsAsFactors = FALSE,
    check.names = FALSE
  )
  
  # Remove the old result for the current seed
  existing_results <- existing_results[
    existing_results$Seed != split_seed,
    ,
    drop = FALSE
  ]
  
  # Existing CSV files may not yet contain the new base-learner
  # columns. Create a common column structure before row-binding.
  all_result_columns <- union(
    colnames(existing_results),
    colnames(seed_result)
  )
  
  columns_missing_from_existing <- setdiff(
    all_result_columns,
    colnames(existing_results)
  )
  
  for (
    column_name in columns_missing_from_existing
  ) {
    
    existing_results[[column_name]] <- rep(
      NA,
      nrow(existing_results)
    )
  }
  
  columns_missing_from_new_result <- setdiff(
    all_result_columns,
    colnames(seed_result)
  )
  
  for (
    column_name in columns_missing_from_new_result
  ) {
    
    seed_result[[column_name]] <- rep(
      NA,
      nrow(seed_result)
    )
  }
  
  existing_results <- existing_results[
    ,
    all_result_columns,
    drop = FALSE
  ]
  
  seed_result <- seed_result[
    ,
    all_result_columns,
    drop = FALSE
  ]
  
  combined_results <- rbind(
    existing_results,
    seed_result
  )
  
} else {
  
  combined_results <- seed_result
}

combined_results <- combined_results[
  order(combined_results$Seed),
  ,
  drop = FALSE
]

rownames(combined_results) <- NULL

write.csv(
  combined_results,
  file = result_filename,
  row.names = FALSE
)

cat(
  "\nResults saved to:\n",
  normalizePath(
    result_filename,
    winslash = "/",
    mustWork = TRUE
  ),
  "\n"
)

combined_results
```


```{r}
# ------------------------------------------------------------
# Save ensemble comparison plots
# ------------------------------------------------------------

plot_directory <- file.path(
  results_directory,
  "plots",
  configuration_id
)

dir.create(
  plot_directory,
  showWarnings = FALSE,
  recursive = TRUE
)

# ============================================================
# Plot 1: balanced accuracy across seeds
# ============================================================

model_plot_data <- combined_results |>
  select(
    Seed,
    XGBoost_BA,
    Hard_Vote_BA,
    Standard_Stack_BA,
    Prefilter_Stack_BA
  ) |>
  pivot_longer(
    cols = -Seed,
    names_to = "Model",
    values_to = "Balanced_Accuracy"
  ) |>
  mutate(
    Model = recode(
      Model,
      XGBoost_BA =
        "Standalone XGBoost",
      Hard_Vote_BA =
        "Hard voting",
      Standard_Stack_BA =
        "Standard stack",
      Prefilter_Stack_BA =
        "Pre-filtered stack"
    )
  )

performance_plot <- ggplot(
  model_plot_data,
  aes(
    x = factor(Seed),
    y = Balanced_Accuracy,
    group = Model,
    shape = Model,
    linetype = Model
  )
) +
  geom_hline(
    yintercept = 0.5,
    linetype = "dotted"
  ) +
  geom_line(
    linewidth = 0.7
  ) +
  geom_point(
    size = 2.5
  ) +
  labs(
    title = paste(
      "Balanced accuracy across seeds:",
      configuration_id
    ),
    subtitle = paste(
      "The dotted reference line represents",
      "balanced accuracy = 0.50"
    ),
    x = "Split seed",
    y = "Balanced accuracy",
    shape = "Method",
    linetype = "Method"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )

performance_plot_filename <- file.path(
  plot_directory,
  "01_balanced_accuracy_across_seeds.png"
)

ggsave(
  filename = performance_plot_filename,
  plot = performance_plot,
  width = 10,
  height = 6,
  units = "in",
  dpi = 300
)

# ============================================================
# Plot 2: pre-filtering difference across seeds
# ============================================================

difference_plot_data <- combined_results |>
  select(
    Seed,
    Prefilter_Minus_Standard,
    Prefilter_Minus_XGBoost
  ) |>
  pivot_longer(
    cols = -Seed,
    names_to = "Comparison",
    values_to = "Balanced_Accuracy_Difference"
  ) |>
  mutate(
    Comparison = recode(
      Comparison,
      Prefilter_Minus_Standard =
        "Pre-filtered minus standard stack",
      Prefilter_Minus_XGBoost =
        "Pre-filtered minus standalone XGBoost"
    )
  )

difference_plot <- ggplot(
  difference_plot_data,
  aes(
    x = factor(Seed),
    y = Balanced_Accuracy_Difference,
    fill = Comparison
  )
) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.7
  ) +
  geom_col(
    position = position_dodge(
      width = 0.8
    ),
    width = 0.7
  ) +
  labs(
    title = "Effect of pre-filtering across seeds",
    subtitle = paste(
      "Positive values favour the pre-filtered ensemble;",
      "negative values favour the comparison method"
    ),
    x = "Split seed",
    y = "Difference in balanced accuracy",
    fill = "Comparison"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )

difference_plot_filename <- file.path(
  plot_directory,
  "02_prefilter_performance_difference.png"
)

ggsave(
  filename = difference_plot_filename,
  plot = difference_plot,
  width = 10,
  height = 6,
  units = "in",
  dpi = 300
)

# ============================================================
# Plot 3: test-case routing across seeds
# ============================================================

routing_plot_data <- combined_results |>
  transmute(
    Seed = factor(Seed),
    `Direct vote` = Direct_Vote_Rate_Test,
    `Meta-learner` = 1 - Direct_Vote_Rate_Test
  ) |>
  pivot_longer(
    cols = c(
      `Direct vote`,
      `Meta-learner`
    ),
    names_to = "Route",
    values_to = "Proportion"
  )

routing_plot <- ggplot(
  routing_plot_data,
  aes(
    x = Seed,
    y = Proportion,
    fill = Route
  )
) +
  geom_col(
    width = 0.75
  ) +
  geom_text(
    aes(
      label = scales::percent(
        Proportion,
        accuracy = 1
      )
    ),
    position = position_stack(
      vjust = 0.5
    ),
    size = 3.5
  ) +
  scale_y_continuous(
    labels = scales::percent_format(
      accuracy = 1
    ),
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  labs(
    title = "Routing of test observations across seeds",
    subtitle = paste(
      "Direct-vote cases bypass XGBoost;",
      "the remaining cases are sent to the meta-learner"
    ),
    x = "Split seed",
    y = "Percentage of test observations",
    fill = "Prediction route"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.major.x = element_blank()
  )

routing_plot_filename <- file.path(
  plot_directory,
  "03_test_case_routing.png"
)

ggsave(
  filename = routing_plot_filename,
  plot = routing_plot,
  width = 10,
  height = 6,
  units = "in",
  dpi = 300
)

# Display plots in the rendered HTML
print(performance_plot)
print(difference_plot)
print(routing_plot)

cat(
  "\nPlots saved to:\n",
  normalizePath(
    plot_directory,
    winslash = "/",
    mustWork = TRUE
  ),
  "\n"
)
```

```{r}
# ------------------------------------------------------------
# Diagnose base-learner and pre-filtered-stack performance
# ------------------------------------------------------------

diagnostic_results <- read.csv(
  result_filename,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

required_base_metrics <- c(
  "Threshold",
  "Accuracy",
  "Balanced_Accuracy",
  "Precision",
  "Recall",
  "Specificity",
  "F1"
)

required_ensemble_columns <- c(
  "Seed",
  "Prefilter_Stack_BA",
  "Standard_Stack_BA",
  "XGBoost_BA",
  "Direct_Vote_Rate_Test",
  "Test_Cases_Sent_To_Meta"
)

missing_ensemble_columns <- setdiff(
  required_ensemble_columns,
  colnames(diagnostic_results)
)

if (length(missing_ensemble_columns) > 0L) {
  stop(
    paste(
      "The result CSV is missing:",
      paste(
        missing_ensemble_columns,
        collapse = ", "
      )
    )
  )
}

# ------------------------------------------------------------
# Convert wide base-learner columns into a long table
# ------------------------------------------------------------

base_learner_long <- diagnostic_results |>
  select(
    Seed,
    starts_with("Base_")
  ) |>
  pivot_longer(
    cols = starts_with("Base_"),
    names_to = c(
      "Model",
      "Metric"
    ),
    names_pattern = paste0(
      "^Base_(.+?)_(",
      paste(
        c(
          "Balanced_Accuracy",
          "Specificity",
          "Precision",
          "Threshold",
          "Accuracy",
          "Recall",
          "F1"
        ),
        collapse = "|"
      ),
      ")$"
    ),
    values_to = "Value"
  ) |>
  pivot_wider(
    names_from = Metric,
    values_from = Value
  ) |>
  arrange(
    Seed,
    Model
  )

# Check that all expected metrics were reconstructed
missing_reconstructed_metrics <- setdiff(
  required_base_metrics,
  colnames(base_learner_long)
)

if (length(missing_reconstructed_metrics) > 0L) {
  stop(
    paste(
      "Could not reconstruct these base-learner metrics:",
      paste(
        missing_reconstructed_metrics,
        collapse = ", "
      )
    )
  )
}

# ------------------------------------------------------------
# Table 1: average base-learner performance across seeds
# ------------------------------------------------------------

base_learner_summary <- base_learner_long |>
  group_by(Model) |>
  summarise(
    Seeds = sum(
      !is.na(Balanced_Accuracy)
    ),
    Mean_BA = mean(
      Balanced_Accuracy,
      na.rm = TRUE
    ),
    SD_BA = sd(
      Balanced_Accuracy,
      na.rm = TRUE
    ),
    Minimum_BA = min(
      Balanced_Accuracy,
      na.rm = TRUE
    ),
    Maximum_BA = max(
      Balanced_Accuracy,
      na.rm = TRUE
    ),
    Mean_Accuracy = mean(
      Accuracy,
      na.rm = TRUE
    ),
    Mean_Precision = mean(
      Precision,
      na.rm = TRUE
    ),
    Mean_Recall = mean(
      Recall,
      na.rm = TRUE
    ),
    Mean_Specificity = mean(
      Specificity,
      na.rm = TRUE
    ),
    Mean_F1 = mean(
      F1,
      na.rm = TRUE
    ),
    Mean_Threshold = mean(
      Threshold,
      na.rm = TRUE
    ),
    SD_Threshold = sd(
      Threshold,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  arrange(
    desc(Mean_BA)
  )

knitr::kable(
  base_learner_summary |>
    mutate(
      across(
        where(is.numeric),
        ~ round(.x, 4)
      )
    ),
  caption = paste(
    "Individual base-learner performance across seeds for",
    configuration_id
  )
)

# ------------------------------------------------------------
# Table 2: strongest and weakest learner for each seed
# ------------------------------------------------------------

base_seed_summary <- base_learner_long |>
  filter(
    !is.na(Balanced_Accuracy)
  ) |>
  group_by(Seed) |>
  summarise(
    Mean_Base_BA = mean(
      Balanced_Accuracy
    ),
    SD_Base_BA = sd(
      Balanced_Accuracy
    ),
    Best_Base_Learner =
      Model[
        which.max(
          Balanced_Accuracy
        )
      ],
    Best_Base_BA = max(
      Balanced_Accuracy
    ),
    Worst_Base_Learner =
      Model[
        which.min(
          Balanced_Accuracy
        )
      ],
    Worst_Base_BA = min(
      Balanced_Accuracy
    ),
    Base_BA_Spread =
      Best_Base_BA -
      Worst_Base_BA,
    Base_Learners_Above_050 = sum(
      Balanced_Accuracy > 0.50
    ),
    .groups = "drop"
  )

prefilter_seed_summary <- diagnostic_results |>
  select(
    Seed,
    Prefilter_Stack_BA,
    Standard_Stack_BA,
    XGBoost_BA,
    Direct_Vote_Rate_Test,
    Test_Cases_Sent_To_Meta
  )

base_learners_beaten <- base_learner_long |>
  left_join(
    diagnostic_results |>
      select(
        Seed,
        Prefilter_Stack_BA
      ),
    by = "Seed"
  ) |>
  group_by(Seed) |>
  summarise(
    Base_Learners_Beaten_By_Prefilter = sum(
      Prefilter_Stack_BA >
        Balanced_Accuracy,
      na.rm = TRUE
    ),
    Base_Learners_Matched_By_Prefilter = sum(
      Prefilter_Stack_BA ==
        Balanced_Accuracy,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

seed_diagnostics <- prefilter_seed_summary |>
  left_join(
    base_seed_summary,
    by = "Seed"
  ) |>
  left_join(
    base_learners_beaten,
    by = "Seed"
  ) |>
  mutate(
    Prefilter_Minus_Mean_Base =
      Prefilter_Stack_BA -
      Mean_Base_BA,
    Prefilter_Minus_Best_Base =
      Prefilter_Stack_BA -
      Best_Base_BA,
    Prefilter_Minus_Standard =
      Prefilter_Stack_BA -
      Standard_Stack_BA
  ) |>
  arrange(Seed)

knitr::kable(
  seed_diagnostics |>
    mutate(
      across(
        where(is.numeric),
        ~ round(.x, 4)
      )
    ),
  caption = paste(
    "Per-seed comparison of the pre-filtered stack",
    "with the individual base learners"
  )
)

# ------------------------------------------------------------
# Table 3: which base learner is best most often?
# ------------------------------------------------------------

best_base_learner_counts <- base_seed_summary |>
  count(
    Best_Base_Learner,
    name = "Number_Of_Seeds"
  ) |>
  mutate(
    Proportion_Of_Seeds =
      Number_Of_Seeds /
      sum(Number_Of_Seeds)
  ) |>
  arrange(
    desc(Number_Of_Seeds),
    Best_Base_Learner
  )

knitr::kable(
  best_base_learner_counts |>
    mutate(
      Proportion_Of_Seeds = round(
        Proportion_Of_Seeds,
        3
      )
    ),
  caption = "Frequency with which each base learner achieved the highest balanced accuracy"
)

# ------------------------------------------------------------
# Safe correlation function
# ------------------------------------------------------------

safe_correlation <- function(x, y) {
  
  complete_index <- complete.cases(
    x,
    y
  )
  
  x_complete <- x[complete_index]
  y_complete <- y[complete_index]
  
  if (
    length(x_complete) < 3L ||
    sd(x_complete) == 0 ||
    sd(y_complete) == 0
  ) {
    return(NA_real_)
  }
  
  cor(
    x_complete,
    y_complete
  )
}

# ------------------------------------------------------------
# Table 4: association between each base learner and
# Prefilter_Stack_BA across seeds
# ------------------------------------------------------------

base_prefilter_association <- base_learner_long |>
  left_join(
    diagnostic_results |>
      select(
        Seed,
        Prefilter_Stack_BA
      ),
    by = "Seed"
  ) |>
  group_by(Model) |>
  summarise(
    Correlation_With_Prefilter_BA =
      safe_correlation(
        Balanced_Accuracy,
        Prefilter_Stack_BA
      ),
    Mean_Base_BA = mean(
      Balanced_Accuracy,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  arrange(
    desc(
      Correlation_With_Prefilter_BA
    )
  )

knitr::kable(
  base_prefilter_association |>
    mutate(
      across(
        where(is.numeric),
        ~ round(.x, 4)
      )
    ),
  caption = paste(
    "Exploratory correlation between base-learner",
    "balanced accuracy and pre-filtered-stack balanced accuracy"
  )
)

# ------------------------------------------------------------
# Table 5: relationship between routing and performance
# ------------------------------------------------------------

routing_association <- data.frame(
  Attribute = c(
    "Direct vote rate",
    "Cases sent to meta-learner",
    "Mean base-learner BA",
    "Best base-learner BA",
    "Spread between best and worst base learner"
  ),
  Correlation_With_Prefilter_BA = c(
    safe_correlation(
      seed_diagnostics$Direct_Vote_Rate_Test,
      seed_diagnostics$Prefilter_Stack_BA
    ),
    safe_correlation(
      seed_diagnostics$Test_Cases_Sent_To_Meta,
      seed_diagnostics$Prefilter_Stack_BA
    ),
    safe_correlation(
      seed_diagnostics$Mean_Base_BA,
      seed_diagnostics$Prefilter_Stack_BA
    ),
    safe_correlation(
      seed_diagnostics$Best_Base_BA,
      seed_diagnostics$Prefilter_Stack_BA
    ),
    safe_correlation(
      seed_diagnostics$Base_BA_Spread,
      seed_diagnostics$Prefilter_Stack_BA
    )
  )
)

knitr::kable(
  routing_association |>
    mutate(
      Correlation_With_Prefilter_BA = round(
        Correlation_With_Prefilter_BA,
        4
      )
    ),
  caption = paste(
    "Exploratory associations with",
    "Prefilter_Stack_BA across seeds"
  )
)

# ------------------------------------------------------------
# Table 6: overall diagnostic summary
# ------------------------------------------------------------

overall_prefilter_diagnosis <- seed_diagnostics |>
  summarise(
    Number_Of_Seeds = n(),
    
    Mean_Prefilter_BA = mean(
      Prefilter_Stack_BA,
      na.rm = TRUE
    ),
    
    SD_Prefilter_BA = sd(
      Prefilter_Stack_BA,
      na.rm = TRUE
    ),
    
    Mean_Of_Mean_Base_BA = mean(
      Mean_Base_BA,
      na.rm = TRUE
    ),
    
    Mean_Best_Base_BA = mean(
      Best_Base_BA,
      na.rm = TRUE
    ),
    
    Mean_Prefilter_Minus_Mean_Base = mean(
      Prefilter_Minus_Mean_Base,
      na.rm = TRUE
    ),
    
    Mean_Prefilter_Minus_Best_Base = mean(
      Prefilter_Minus_Best_Base,
      na.rm = TRUE
    ),
    
    Seeds_Prefilter_Beats_Mean_Base = sum(
      Prefilter_Minus_Mean_Base > 0,
      na.rm = TRUE
    ),
    
    Seeds_Prefilter_Beats_Best_Base = sum(
      Prefilter_Minus_Best_Base > 0,
      na.rm = TRUE
    ),
    
    Seeds_Prefilter_Beats_Standard_Stack = sum(
      Prefilter_Minus_Standard > 0,
      na.rm = TRUE
    ),
    
    Mean_Base_Learners_Beaten = mean(
      Base_Learners_Beaten_By_Prefilter,
      na.rm = TRUE
    )
  )

knitr::kable(
  overall_prefilter_diagnosis |>
    mutate(
      across(
        where(is.numeric),
        ~ round(.x, 4)
      )
    ),
  caption = paste(
    "Overall diagnosis of",
    "Prefilter_Stack_BA"
  )
)

# ------------------------------------------------------------
# Consistency check:
# standalone XGBoost versus XGBoost stored as a base learner
# ------------------------------------------------------------

if (
  "Base_XGBoost_Balanced_Accuracy" %in%
    colnames(diagnostic_results)
) {
  
  xgboost_consistency_check <- data.frame(
    Maximum_Absolute_Difference = max(
      abs(
        diagnostic_results$XGBoost_BA -
          diagnostic_results$
            Base_XGBoost_Balanced_Accuracy
      ),
      na.rm = TRUE
    )
  )
  
  knitr::kable(
    xgboost_consistency_check |>
      mutate(
        Maximum_Absolute_Difference = round(
          Maximum_Absolute_Difference,
          8
        )
      ),
    caption = paste(
      "Consistency check between standalone XGBoost",
      "and XGBoost recorded as a base learner"
    )
  )
}
```

```{r}
# ------------------------------------------------------------
# Save test-level predictions and vote patterns
# ------------------------------------------------------------

model_names <- colnames(
  stack_test_features
)

# Retrieve the validation-selected threshold for each learner
base_threshold_lookup <- setNames(
  base_learner_test_results$Threshold,
  base_learner_test_results$Model
)

missing_thresholds <- setdiff(
  model_names,
  names(base_threshold_lookup)
)

if (length(missing_thresholds) > 0L) {
  stop(
    paste(
      "Missing validation-selected thresholds for:",
      paste(
        missing_thresholds,
        collapse = ", "
      )
    )
  )
}

# ------------------------------------------------------------
# Base-learner predictions using validation-selected thresholds
# ------------------------------------------------------------

base_selected_predictions <- sapply(
  model_names,
  function(model_name) {
    
    as.integer(
      stack_test_features[[model_name]] >=
        base_threshold_lookup[[model_name]]
    )
  }
)

colnames(base_selected_predictions) <- paste0(
  "Prediction_",
  model_names
)

base_selected_predictions <- as.data.frame(
  base_selected_predictions,
  check.names = FALSE
)

# ------------------------------------------------------------
# Base-learner gate votes using learner-specific OOF thresholds
# ------------------------------------------------------------

paper_vote_predictions <- as.data.frame(
  paper_test_votes$hard_predictions * 1L,
  check.names = FALSE
)

colnames(paper_vote_predictions) <- paste0(
  "GateVoteOOF_",
  model_names
)

# ------------------------------------------------------------
# Store base-learner probabilities
# ------------------------------------------------------------

base_probability_results <- as.data.frame(
  stack_test_features,
  check.names = FALSE
)

colnames(base_probability_results) <- paste0(
  "Probability_",
  model_names
)

# ------------------------------------------------------------
# Recover original dataset row numbers
# ------------------------------------------------------------

test_row_id <- suppressWarnings(
  as.integer(
    rownames(test)
  )
)

if (anyNA(test_row_id)) {
  test_row_id <- seq_len(
    nrow(test)
  )
}

actual_test_class <- as.integer(
  as.character(
    test$spam
  )
)

# ------------------------------------------------------------
# Construct one row per test observation
# ------------------------------------------------------------

test_prediction_results <- cbind(
  data.frame(
    Seed = split_seed,
    Configuration_ID = configuration_id,
    Test_Row_ID = test_row_id,
    Actual_Class = actual_test_class,
    stringsAsFactors = FALSE
  ),
  
  base_probability_results,
  base_selected_predictions,
  paper_vote_predictions,
  
  data.frame(
    Positive_Votes =
      paper_test_votes$summary$Positive_Votes,
    
    Negative_Votes =
      paper_test_votes$summary$Negative_Votes,
    
    Maximum_Votes =
      paper_test_votes$summary$Maximum_Votes,
    
    Majority_Label =
      paper_test_votes$summary$Majority_Label,
    
    Confident_Vote =
      paper_test_votes$summary$Confident,
    
    Prediction_Route = ifelse(
      paper_test_votes$summary$Confident,
      "Direct_vote",
      "Meta_learner"
    ),
    
    Paper_Meta_Probability =
      paper_test_meta_prob,
    
    Paper_Meta_Only_Prediction = as.integer(
      paper_test_meta_prob >=
        paper_meta_threshold
    ),
    
    Prefilter_Final_Prediction = as.integer(
      as.character(
        paper_prefilter_pred
      )
    ),
    
    Standard_Stack_Probability =
      stack_prob,
    
    Standard_Stack_Prediction = as.integer(
      as.character(
        stack_pred
      )
    ),
    
    Standalone_XGBoost_Probability =
      xgb_prob,
    
    Standalone_XGBoost_Prediction = as.integer(
      as.character(
        xgb_pred
      )
    ),
    
    Hard_Vote_Prediction = as.integer(
      as.character(
        hard_vote_pred
      )
    ),
    
    stringsAsFactors = FALSE
  )
)

# Correctness indicators
test_prediction_results$Prefilter_Correct <-
  as.integer(
    test_prediction_results$
      Prefilter_Final_Prediction ==
      test_prediction_results$Actual_Class
  )

test_prediction_results$Standard_Stack_Correct <-
  as.integer(
    test_prediction_results$
      Standard_Stack_Prediction ==
      test_prediction_results$Actual_Class
  )

test_prediction_results$Standalone_XGBoost_Correct <-
  as.integer(
    test_prediction_results$
      Standalone_XGBoost_Prediction ==
      test_prediction_results$Actual_Class
  )

test_prediction_results$Hard_Vote_Correct <-
  as.integer(
    test_prediction_results$
      Hard_Vote_Prediction ==
      test_prediction_results$Actual_Class
  )

# ------------------------------------------------------------
# Helper: replace the current seed and append the new rows
# ------------------------------------------------------------

replace_seed_and_save <- function(
    new_results,
    filename,
    seed_value) {
  
  if (file.exists(filename)) {
    
    existing_results <- read.csv(
      filename,
      stringsAsFactors = FALSE,
      check.names = FALSE
    )
    
    existing_results <- existing_results[
      existing_results$Seed != seed_value,
      ,
      drop = FALSE
    ]
    
    all_columns <- union(
      colnames(existing_results),
      colnames(new_results)
    )
    
    for (
      column_name in setdiff(
        all_columns,
        colnames(existing_results)
      )
    ) {
      existing_results[[column_name]] <- NA
    }
    
    for (
      column_name in setdiff(
        all_columns,
        colnames(new_results)
      )
    ) {
      new_results[[column_name]] <- NA
    }
    
    existing_results <- existing_results[
      ,
      all_columns,
      drop = FALSE
    ]
    
    new_results <- new_results[
      ,
      all_columns,
      drop = FALSE
    ]
    
    combined_results <- rbind(
      existing_results,
      new_results
    )
    
  } else {
    
    combined_results <- new_results
  }
  
  combined_results <- combined_results[
    order(
      combined_results$Seed
    ),
    ,
    drop = FALSE
  ]
  
  rownames(combined_results) <- NULL
  
  write.csv(
    combined_results,
    filename,
    row.names = FALSE
  )
  
  invisible(
    combined_results
  )
}

# ------------------------------------------------------------
# Save observation-level predictions
# ------------------------------------------------------------

test_prediction_filename <- file.path(
  results_directory,
  paste0(
    configuration_id,
    "_test_predictions.csv"
  )
)

all_test_prediction_results <- replace_seed_and_save(
  new_results = test_prediction_results,
  filename = test_prediction_filename,
  seed_value = split_seed
)

# ------------------------------------------------------------
# Pairwise prediction-level diversity
# ------------------------------------------------------------

base_prediction_matrix <- as.matrix(
  base_selected_predictions
)

colnames(base_prediction_matrix) <- model_names

model_pairs <- combn(
  model_names,
  2,
  simplify = FALSE
)

pairwise_diversity_results <- lapply(
  model_pairs,
  function(model_pair) {
    
    model_1 <- model_pair[1]
    model_2 <- model_pair[2]
    
    prediction_1 <-
      base_prediction_matrix[, model_1]
    
    prediction_2 <-
      base_prediction_matrix[, model_2]
    
    correct_1 <-
      prediction_1 == actual_test_class
    
    correct_2 <-
      prediction_2 == actual_test_class
    
    both_correct <- sum(
      correct_1 & correct_2
    )
    
    model_1_only_correct <- sum(
      correct_1 & !correct_2
    )
    
    model_2_only_correct <- sum(
      !correct_1 & correct_2
    )
    
    both_wrong <- sum(
      !correct_1 & !correct_2
    )
    
    q_denominator <-
      both_correct * both_wrong +
      model_1_only_correct *
      model_2_only_correct
    
    q_statistic <- ifelse(
      q_denominator == 0,
      NA_real_,
      (
        both_correct * both_wrong -
          model_1_only_correct *
          model_2_only_correct
      ) /
        q_denominator
    )
    
    data.frame(
      Seed = split_seed,
      Configuration_ID = configuration_id,
      Model_1 = model_1,
      Model_2 = model_2,
      
      Disagreement_Rate = mean(
        prediction_1 != prediction_2
      ),
      
      Double_Fault_Rate = mean(
        !correct_1 & !correct_2
      ),
      
      Both_Correct_Rate = mean(
        correct_1 & correct_2
      ),
      
      Model_1_Only_Correct_Rate = mean(
        correct_1 & !correct_2
      ),
      
      Model_2_Only_Correct_Rate = mean(
        !correct_1 & correct_2
      ),
      
      Q_Statistic = q_statistic,
      
      stringsAsFactors = FALSE
    )
  }
)

pairwise_diversity_results <- do.call(
  rbind,
  pairwise_diversity_results
)

pairwise_diversity_filename <- file.path(
  results_directory,
  paste0(
    configuration_id,
    "_pairwise_diversity.csv"
  )
)

all_pairwise_diversity_results <-
  replace_seed_and_save(
    new_results =
      pairwise_diversity_results,
    filename =
      pairwise_diversity_filename,
    seed_value =
      split_seed
  )

cat(
  "\nTest-level predictions saved to:\n",
  normalizePath(
    test_prediction_filename,
    winslash = "/",
    mustWork = TRUE
  ),
  "\n\nPairwise diversity results saved to:\n",
  normalizePath(
    pairwise_diversity_filename,
    winslash = "/",
    mustWork = TRUE
  ),
  "\n"
)

pairwise_diversity_results
```

```{r visual-diagnostics, fig.width=10, fig.height=6}
# ------------------------------------------------------------
# Visual diagnostics for base learners and prediction diversity
# ------------------------------------------------------------

required_objects <- c(
  "base_learner_long",
  "seed_diagnostics",
  "all_pairwise_diversity_results",
  "all_test_prediction_results"
)

missing_objects <- required_objects[
  !vapply(
    required_objects,
    exists,
    logical(1)
  )
]

if (length(missing_objects) > 0L) {
  stop(
    paste(
      "Run the preceding diagnostic chunks first. Missing objects:",
      paste(missing_objects, collapse = ", ")
    )
  )
}

diagnostic_plot_directory <- file.path(
  results_directory,
  "plots",
  configuration_id,
  "diagnostics"
)

dir.create(
  diagnostic_plot_directory,
  showWarnings = FALSE,
  recursive = TRUE
)

# ------------------------------------------------------------
# Readable model labels
# ------------------------------------------------------------

clean_model_name <- function(model_name) {
  
  recode(
    model_name,
    ElasticNet = "Elastic Net",
    GAM = "GAM",
    MARS = "MARS",
    RandomForest = "Random Forest",
    XGBoost = "XGBoost",
    SVM = "SVM",
    .default = model_name
  )
}

save_diagnostic_plot <- function(
    plot_object,
    filename,
    width = 10,
    height = 6) {
  
  ggsave(
    filename = file.path(
      diagnostic_plot_directory,
      filename
    ),
    plot = plot_object,
    width = width,
    height = height,
    units = "in",
    dpi = 300
  )
}

# ============================================================
# Plot 1: distribution of base-learner balanced accuracy
# ============================================================

base_plot_data <- base_learner_long |>
  filter(
    !is.na(Balanced_Accuracy)
  ) |>
  mutate(
    Model = clean_model_name(Model)
  )

base_model_order <- base_plot_data |>
  group_by(Model) |>
  summarise(
    Mean_BA = mean(
      Balanced_Accuracy,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  arrange(Mean_BA) |>
  pull(Model)

base_plot_data <- base_plot_data |>
  mutate(
    Model = factor(
      Model,
      levels = base_model_order
    )
  )

base_distribution_plot <- ggplot(
  base_plot_data,
  aes(
    x = Model,
    y = Balanced_Accuracy
  )
) +
  geom_hline(
    yintercept = 0.5,
    linetype = "dashed"
  ) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA
  ) +
  geom_jitter(
    width = 0.08,
    height = 0,
    size = 2,
    alpha = 0.75
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 3.5
  ) +
  coord_flip() +
  labs(
    title = "Base-learner performance across seeds",
    subtitle = paste(
      "Each circle is one split seed;",
      "diamonds indicate the mean"
    ),
    x = NULL,
    y = "Balanced accuracy"
  ) +
  theme_minimal()

save_diagnostic_plot(
  base_distribution_plot,
  "04_base_learner_balanced_accuracy.png"
)

print(base_distribution_plot)

# ============================================================
# Plot 2: base-learner performance heatmap
# ============================================================

base_heatmap_plot <- ggplot(
  base_plot_data,
  aes(
    x = factor(Seed),
    y = Model,
    fill = Balanced_Accuracy
  )
) +
  geom_tile(
    linewidth = 0.5
  ) +
  geom_text(
    aes(
      label = sprintf(
        "%.3f",
        Balanced_Accuracy
      )
    ),
    size = 3.2
  ) +
  scale_fill_gradient2(
    midpoint = 0.5,
    labels = scales::number_format(
      accuracy = 0.01
    )
  ) +
  labs(
    title = "Balanced accuracy by learner and split seed",
    subtitle = paste(
      "Values above 0.50 indicate performance",
      "above the balanced no-skill reference"
    ),
    x = "Split seed",
    y = NULL,
    fill = "Balanced\naccuracy"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    legend.position = "right"
  )

save_diagnostic_plot(
  base_heatmap_plot,
  "05_base_learner_seed_heatmap.png"
)

print(base_heatmap_plot)

# ============================================================
# Plot 3: pre-filtered stack versus individual learners
# ============================================================

prefilter_base_comparison <- seed_diagnostics |>
  select(
    Seed,
    Prefilter_Minus_Mean_Base,
    Prefilter_Minus_Best_Base
  ) |>
  pivot_longer(
    cols = -Seed,
    names_to = "Comparison",
    values_to = "Balanced_Accuracy_Difference"
  ) |>
  mutate(
    Comparison = recode(
      Comparison,
      Prefilter_Minus_Mean_Base =
        "Pre-filtered minus average base learner",
      Prefilter_Minus_Best_Base =
        "Pre-filtered minus best base learner"
    )
  )

prefilter_base_plot <- ggplot(
  prefilter_base_comparison,
  aes(
    x = factor(Seed),
    y = Balanced_Accuracy_Difference,
    fill = Comparison
  )
) +
  geom_hline(
    yintercept = 0,
    linewidth = 0.7
  ) +
  geom_col(
    position = position_dodge(
      width = 0.8
    ),
    width = 0.7
  ) +
  labs(
    title = "Does the pre-filtered stack add value?",
    subtitle = paste(
      "Positive bars favour the pre-filtered stack;",
      "negative bars favour the base learners"
    ),
    x = "Split seed",
    y = "Difference in balanced accuracy",
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )

save_diagnostic_plot(
  prefilter_base_plot,
  "06_prefilter_vs_base_learners.png"
)

print(prefilter_base_plot)

# ============================================================
# Prepare pairwise diversity summaries
# ============================================================

pairwise_summary <- all_pairwise_diversity_results |>
  filter(
    Configuration_ID == configuration_id
  ) |>
  group_by(
    Model_1,
    Model_2
  ) |>
  summarise(
    Mean_Disagreement = mean(
      Disagreement_Rate,
      na.rm = TRUE
    ),
    Mean_Double_Fault = mean(
      Double_Fault_Rate,
      na.rm = TRUE
    ),
    Mean_Q = mean(
      Q_Statistic,
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(
    Model_1 = clean_model_name(Model_1),
    Model_2 = clean_model_name(Model_2)
  )

all_base_models <- sort(
  unique(
    c(
      pairwise_summary$Model_1,
      pairwise_summary$Model_2
    )
  )
)

make_symmetric_pairwise <- function(
    pairwise_data,
    value_column) {
  
  forward <- pairwise_data |>
    transmute(
      Model_1,
      Model_2,
      Value = .data[[value_column]]
    )
  
  reverse <- forward |>
    transmute(
      Model_1 = Model_2,
      Model_2 = Model_1,
      Value
    )
  
  bind_rows(
    forward,
    reverse
  ) |>
    mutate(
      Model_1 = factor(
        Model_1,
        levels = all_base_models
      ),
      Model_2 = factor(
        Model_2,
        levels = rev(all_base_models)
      )
    )
}

# ============================================================
# Plot 4: pairwise disagreement heatmap
# ============================================================

disagreement_heatmap_data <- make_symmetric_pairwise(
  pairwise_summary,
  "Mean_Disagreement"
)

disagreement_heatmap_plot <- ggplot(
  disagreement_heatmap_data,
  aes(
    x = Model_1,
    y = Model_2,
    fill = Value
  )
) +
  geom_tile(
    linewidth = 0.5
  ) +
  geom_text(
    aes(
      label = scales::percent(
        Value,
        accuracy = 1
      )
    ),
    size = 3.2
  ) +
  scale_fill_gradient(
    labels = scales::percent_format(
      accuracy = 1
    )
  ) +
  coord_equal() +
  labs(
    title = "How often do the base learners disagree?",
    subtitle = paste(
      "Higher disagreement indicates greater prediction diversity,",
      "but does not by itself imply better performance"
    ),
    x = NULL,
    y = NULL,
    fill = "Disagreement"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 35,
      hjust = 1
    )
  )

save_diagnostic_plot(
  disagreement_heatmap_plot,
  "07_pairwise_disagreement_heatmap.png",
  width = 9,
  height = 7
)

print(disagreement_heatmap_plot)

# ============================================================
# Plot 5: pairwise double-fault heatmap
# ============================================================

double_fault_heatmap_data <- make_symmetric_pairwise(
  pairwise_summary,
  "Mean_Double_Fault"
)

double_fault_heatmap_plot <- ggplot(
  double_fault_heatmap_data,
  aes(
    x = Model_1,
    y = Model_2,
    fill = Value
  )
) +
  geom_tile(
    linewidth = 0.5
  ) +
  geom_text(
    aes(
      label = scales::percent(
        Value,
        accuracy = 1
      )
    ),
    size = 3.2
  ) +
  scale_fill_gradient(
    labels = scales::percent_format(
      accuracy = 1
    )
  ) +
  coord_equal() +
  labs(
    title = "How often are two base learners wrong together?",
    subtitle = paste(
      "Lower double-fault rates are preferable;",
      "high values indicate correlated errors"
    ),
    x = NULL,
    y = NULL,
    fill = "Double-fault\nrate"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(
      angle = 35,
      hjust = 1
    )
  )

save_diagnostic_plot(
  double_fault_heatmap_plot,
  "08_pairwise_double_fault_heatmap.png",
  width = 9,
  height = 7
)

print(double_fault_heatmap_plot)

# ============================================================
# Plot 6: pre-filtered performance by prediction route
# ============================================================

route_balanced_accuracy <- function(
    actual,
    predicted) {
  
  actual <- as.integer(actual)
  predicted <- as.integer(predicted)
  
  if (
    !all(c(0, 1) %in% unique(actual))
  ) {
    return(NA_real_)
  }
  
  recall <- mean(
    predicted[actual == 1] == 1
  )
  
  specificity <- mean(
    predicted[actual == 0] == 0
  )
  
  mean(
    c(
      recall,
      specificity
    )
  )
}

route_performance <- all_test_prediction_results |>
  filter(
    Configuration_ID == configuration_id
  ) |>
  mutate(
    Prediction_Route = recode(
      Prediction_Route,
      Direct_vote = "Direct vote",
      Meta_learner = "Meta-learner"
    )
  ) |>
  group_by(
    Seed,
    Prediction_Route
  ) |>
  summarise(
    Cases = n(),
    Balanced_Accuracy = route_balanced_accuracy(
      Actual_Class,
      Prefilter_Final_Prediction
    ),
    .groups = "drop"
  )

route_performance_plot <- ggplot(
  route_performance,
  aes(
    x = factor(Seed),
    y = Balanced_Accuracy,
    fill = Prediction_Route
  )
) +
  geom_hline(
    yintercept = 0.5,
    linetype = "dashed"
  ) +
  geom_col(
    position = position_dodge(
      width = 0.8
    ),
    width = 0.7,
    na.rm = TRUE
  ) +
  geom_text(
    aes(
      label = ifelse(
        is.na(Balanced_Accuracy),
        paste0(
          "N/A\nn=",
          Cases
        ),
        paste0(
          sprintf(
            "%.3f",
            Balanced_Accuracy
          ),
          "\nn=",
          Cases
        )
      )
    ),
    position = position_dodge(
      width = 0.8
    ),
    vjust = -0.25,
    size = 3,
    na.rm = FALSE
  ) +
  coord_cartesian(
    ylim = c(0, 1.08)
  ) +
  labs(
    title = "Pre-filtered ensemble performance by prediction route",
    subtitle = paste(
      "This separates cases handled directly by voting",
      "from cases handled by the XGBoost meta-learner"
    ),
    x = "Split seed",
    y = "Balanced accuracy within route",
    fill = "Prediction route",
    caption = paste(
      "N/A indicates that the route contained only one true class,",
      "so balanced accuracy was not defined."
    )
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )

save_diagnostic_plot(
  route_performance_plot,
  "09_prefilter_performance_by_route.png"
)

print(route_performance_plot)

cat(
  "\nDiagnostic plots saved to:\n",
  normalizePath(
    diagnostic_plot_directory,
    winslash = "/",
    mustWork = TRUE
  ),
  "\n"
)
```


